Building the Million Transcriptome Ocular Meta-Atlas

### 2021-01-16 manuscript_00.Rmd
text update

scPOP is a lightweight, low dependency R package which brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and re-implements the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms. The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better). NMI and ARI were used to assess the consistency of cell type to cluster assignment (where 1 is perfect correspondence between cluster and cell type).

@@

We see in the across the sumZScale metric, which we define as sumZScale = zscale(LISI batch) + zscale(LISI cluster) + zscale(LISI cell type) + zscale(silhouette batch) + zscale(silhouette cluster) + zscale(silhouette cell type) + zscale(NMI) + zscale(ARI) that scVI overall performs best. As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c). We see that the remaining normalization approaches alters the batch correction performance, though the exact outcome differs across the different methods. We also see that changing the number of latent dimensions (8 or 30) can occasionally dramatically change performance. These results demonstrate the importance of assessing performance in a rigorous manner aceoss many parameters.

@@

@@

Projection of new data onto scEiaD potentially allows for semi-direct comparison of other species, *in vitro* systems, and pharamacologic / mutational perturbation of mammalian retina cells.

We demonstrate with a provided iPSC-based RPE cell line that we sequenced that non-tissue can be projected onto the scEiaD

>>>>

scPOP is a lightweight, low dependency R package which brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and brings in the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms from aricode. The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better). NMI and ARI were used to assess the consistency of cell type to cluster assignment (where 1 is perfect correspondence between cluster and cell type).

@@

To merge the different scores we define $sumZScale = \sum scale(m)$ where m is a metric (LISI by batch, LISI by cluster, LISI by cell type, silhouette by batch, silhouette by cluster, silhouette by celltype, NMI, and ARI). In this scaled metric that weights the different benchmarking algorithms evently, scVI performs best. As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c). We see that the remaining normalization approaches alters the batch correction performance, though the exact outcome differs across the different methods. We also see that changing the number of latent dimensions (8 or 30) can occasionally dramatically change performance. These results demonstrate the importance of assessing performance in a rigorous manner aceoss many parameters.

@@

## Projection of new data onto scEiaD shows that outside datasets for both tissue and iPSC-based samples overlap the scEiaD/scVI model

@@

Projection of new data onto scEiaD allows for semi-direct comparison of other species, *in vitro* systems, and pharamacologic / mutational perturbation of mammalian retina cells. In the previous example we demonstrated how a outside mouse pan-retina dataset maps onto

We demonstrate with a iPSC-based RPE cell line that we ran with the 10x v2 system that non-primary cells can be projected onto the scEiaD. The differentiated iPSC-based RPE from this highly optimized differentiation protocol (CITE BHARTI) overlap post-mortem adult human RPE cells from the Cowan et al. work. (MAKE FIGURE)
